Least-Squares Time Reversal of PDEs

### 2022-07-20 adjop/lsrtm/lsrtm.tex
minor change
taylor series work: paper

\documentclass[longbibliography,twocolumn,amsmath,amssymb,aps,nofootinbib]{revtex4-2}

@@

@@

Time-reversible wave equations of the form
\begin{align}
  \partial_t u &= \sum_{n \geq 1} c_n \partial^{2n - 1}_x u \label{waveeq}
  \intertext{are trivial cases of this optimization problem, because the quantity $\davg{u^2(x, t)}$ is conserved. 
  This implies that the objective functional's levelsets form sphere's in $u'$-space, because its value is given by $\davg{u'^2(T)} = \davg{u'^2(0)}$.
  The spatial derivatives in (\ref{waveeq}) are anti-self-adjoint (i.e. $L[u] = -L^{\dagger}[u]$), so the path of steepest descent is always in the radial derivation }
  \frac{\delta H}{\delta u(x, 0)} &= -u'(x, 0) \nonumber
  \intertext{In contrast, diffusive equations of the form}
  \partial_t u &= \sum_{n \geq 1} b_n \partial^{2n}_x u \label{waveeq}
  \intertext{do not conserve $\davg{u^2(x, t)}$, and consequently their objective functionals depend on $u'(x, 0)$'s \textit{direction} along with its magnitude. Suppose initial deviation can be written as}
  u'(x, 0) &= \sum_{k \geq 0} a_k e^{ikx}.
  \intertext{then}
  \davg{u'(x, T)} &= \pi \sum_{k \geq 0} \Big[\exp\Big({-2T\sum_{n \geq 1} k^{2n}  }\Big)a_k^2\Big].
\end{align}
When the initial deviation $u'(x, 0)$ is described using a basis of fourier coefficients, its levelsets form hyperellipses, whose semi-major (semi-minor) axes are aligned with $u'(x, 0)$'s largest (smallest) fourier coefficient.
Accordingly, the path of steepest descent is only aligned with $u'(x, 0)$ when $u'(x, 0)$ is composed of a single fourier mode. Such systems are self-adjoint, meaning $L[u] = L^{\dagger}[u]$

\begin{figure}
  \includegraphics[width=9cm]{objtest_a0p03b0p0c0p0T3p0R0p015kt10p0kt20.png}
  \caption{Objective function and an example gradient descent path for the diffusion equation. The target simulation $U(t) = 0$ and the deviation is composed of two fourier modes, i.e. $u'(t) = a_1\sin x + a_5\sin 5x$. Because the system is linear self-adjoint, each subsequent guess in the gradient descent trajectory is constrained to lie in this 2D phase space.}
\end{figure}

>>>>

% \documentclass[longbibliography,twocolumn,amsmath,amssymb,aps,nofootinbib]{revtex4-2}
\documentclass[longbibliography,amsmath,amssymb,aps,nofootinbib]{revtex4-2}

@@

% \vspace{1cm}
% \textbf{Grad and Div}

% suppose $f(x, y, z):\mathbb{R}^3 \to \mathbb{R}$

% suppose $\vec{F}(x, y, z):\mathbb{R}^3 \to \mathbb{R}^3$
% \begin{align*}
%   \textbf{Gradient:}&\quad \grad f(x,y,z) \quad \longrightarrow \quad vector \\
%   \textbf{Divergence:}&\quad \grad \cdot \vec{F}(x,y,z) \quad \longrightarrow \quad scalar
% \end{align*}
% \vspace{1cm}

% \textbf{Navier-Stokes Equation}

% \begin{align*}
%   \intertext{Newton's 2nd law:}
%   \vec{F} &= m\vec{a} \\
%   &= \frac{d}{dt} [m\vec{u}]
% \end{align*}

% \begin{align*}
%   \intertext{Consider a velocity $\vec{u}(x, y, t)$}
%   \textbf{Conservation of momentum:} &\quad \partial_t \vec{u} = - \vec{u} \cdot \grad\vec{u} - \grad p + \nu \grad^2 \vec{u} + \vec{F} \tag{1}\\
%   \text{Inertial Term:} &\quad \vec{u}\cdot\grad\vec{u} \\
%   \text{Pressure Gradient Term:} &\quad \grad p \\
%   \text{Diffusive (Friction) Term:} &\quad \nu \laplacian \vec{u} \\
%   \text{Body Force:} &\quad \vec{F} \\
%   &\; \\
%   \textbf{Conservation of mass:} &\quad \grad \cdot \vec{u} = 0 \tag{2}\\
%   p(x, y, t) &= f(t) \; [\sum \sum (a_n\sin(nx)b_m\sin(my))]
% \end{align*}

@@

Time-reversible wave equations of the form
\begin{align}
  \partial_t u &= \sum_{n \geq 1} c_n \partial^{2n - 1}_x u \label{waveeq}
  \intertext{are trivial cases of this optimization problem, because the quantity $\davg{u^2(x, t)}$ is conserved. 
  This implies that the objective functional's levelsets form sphere's in $u'$-space, because its value is given by $\davg{u'^2(T)} = \davg{u'^2(0)}$.
  The spatial derivatives in (\ref{waveeq}) are anti-self-adjoint (i.e. $L[u] = -L^{\dagger}[u]$), so the path of steepest descent is always in the radial derivation }
  \frac{\delta H}{\delta u(x, 0)} &= -u'(x, 0) \nonumber
  \intertext{In contrast, diffusive equations of the form}
  \partial_t u &= \sum_{n \geq 1} b_n \partial^{2n}_x u \label{waveeq}
  \intertext{do not conserve $\davg{u^2(x, t)}$, and consequently their objective functionals depend on $u'(x, 0)$'s \textit{direction} along with its magnitude. Suppose initial deviation can be written as}
  u'(x, 0) &= \sum_{k \geq 0} a_k e^{ikx}.
  \intertext{Let's just consider the diffusion equation ($n=1$). Then the end state is given by}
  u'(x, T) &= \sum_{k\geq 0} a_k \exp(ikx-Tb_1k^2) \label{diffsoln}\\
  \intertext{The nominal objective functional is given by}
  \davg{|u'(x, T)|^2} &= \pi \sum_{k \geq 0} \Big[a_k^2\exp\Big({-2T\sum_{n \geq 1} b_n k^{2n}  }\Big)\, \Big] \label{nomobjdiff}.
  \intertext{Equation (\ref{nomobjdiff}) gives us the elliptical heatmaps we observed previously.}
  \intertext{Let's take a second-order spatial derivative of (\ref{diffsoln})}
  \partial_x^2 u'(x, T) &= \pi \sum_{k\geq 0} -k^2 a_k^2 \exp(ikx-Tb_1k^2)\\
  \intertext{In general, applying a spatial derivative of order $2m$ to (\ref{diffsoln}), yields}
  \partial_x^{2m} u'(x, T) &= \pi \sum_{k\geq 0} (-k^2)^m a_k^2 \exp(ikx-Tb_1k^{2})\\
  \intertext{We can leverage this fact to construct an objective functional whose levelsets are spherical. Consider}
\end{align}
\begin{tiny}
\begin{align*}
  H[u'(T)] &= \davg{|u'(x, T) - (Tb_1)\partial_x^2 u'(x, T) + \frac{(Tb_1)^2}{2!}\partial_x^4 u'(x, T) - \frac{(Tb_1)^3}{3!}\partial_x^6 u'(x, T) ...|^2} \\
  &= \davg{\Big|\sum_{k\geq 0}a_k \exp(ikx-Tb_1k^2) + \sum_{k\geq 0}(Tb_1k^2)a_k \exp(ikx-Tb_1k^2) + \sum_{k\geq 0}\frac{(Tb_1k^2)^2}{2!}a_k \exp(ikx-Tb_1k^2) + \sum_{k\geq 0}\frac{(Tb_1k^2)^3}{3!}a_k \exp(ikx-Tb_1k^2) + ...\Big|^2} \\
  &= \davg{|\sum_{k\geq 0}\Big[a_k \exp(ikx-Tb_1k^2)\Big(1 + (Tb_1k^2) + \frac{(Tb_1k^2)^2}{2!} + \frac{(Tb_1k^2)^3}{3!} ...\Big)\Big]|^2} \\
  &= \davg{|\sum_{k\geq 0}\Big[ a_k \exp(ikx-Tb_1k^2) \Big( \exp(Tb_1k^2) \Big)\Big]|^2} \\
  &= \davg{|\sum_{k\geq 0}\Big[ a_k \exp(ikx)  \Big)\Big]|^2} \\
  &= \pi \sum_{k\geq 0} a_k^2
  % &= \pi \sum_{k \geq 0} \Big[\exp\Big({-2T\sum_{n \geq 1} b_n k^{2n}  }\Big)\, a_k^2\Big].
\end{align*}
\end{tiny}
That's gives us spherical objective levelsets. Notice what happens as we increment the number of terms in the above Taylor Series

% \begin{figure*}
%   \centering
%   \begin{tabular}{@{}c@{}}
%       \includegraphics[width=3.375in]{flux_sup_n.pdf}
%   \end{tabular}
%   \begin{tabular}{@{}c@{}}
%       \includegraphics[width=3.375in]{EV_spectra_2ra.pdf}
%   \end{tabular}
%   \caption{Heat fluxes (left) and eigenvalue spectra (right) of equilibrated states with $\Ra = 2 \times 10^5$ (top) and $\Ra = 10^9$ (bottom). 
%   The heat flux profiles are symmetric about $z=0$, so we only plot for $z<0$.
%   We plot the advection profiles for marginally-stable modes. 
%   At low $\Ra$, a single mode with $k_x = 1.5\pi$ is sufficient to oppose boundary layer diffusion and facilitate heat flux throughout the bulk of the domain. 
%   At large $\Ra$, high-wavenumber modes contribute pronounced small-scale advection profiles which tightly hug the thin boundary layers. 
%   A combination of progressively wider advection profiles is necessary to transition to bulk advection provided by the $k_x = 1.5\pi$ mode. 
%   This bulk mode mimics the large-scale convective cells which are characteristic of Rayleigh–Bénard convection, both in qualitative structure and number of cells in the domain.
%   The eigenfunctions which compose the advective profiles are illustrated in Appendix~\ref{sec:eigenfunctions}.}
%   \label{fig:flux}
% \end{figure*}

\clearpage
When the initial deviation $u'(x, 0)$ is described using a basis of fourier coefficients, its levelsets form hyperellipses, whose semi-major (semi-minor) axes are aligned with $u'(x, 0)$'s largest (smallest) fourier coefficient.
Accordingly, the path of steepest descent is only aligned with $u'(x, 0)$ when $u'(x, 0)$ is composed of a single fourier mode. Such systems are self-adjoint, meaning $L[u] = L^{\dagger}[u]$

% \begin{figure}
%   \includegraphics[width=9cm]{objtest_a0p03b0p0c0p0T3p0R0p015kt10p0kt20.png}
%   \caption{Objective function and an example gradient descent path for the diffusion equation. The target simulation $U(t) = 0$ and the deviation is composed of two fourier modes, i.e. $u'(t) = a_1\sin x + a_5\sin 5x$. Because the system is linear self-adjoint, each subsequent guess in the gradient descent trajectory is constrained to lie in this 2D phase space.}
% \end{figure}
